---
title: 'A7: Clustering and Association Rule Mining'
author: "Jamison Orton"
date: "2026-07-21"
output: 
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 15
    fig_height: 10
editor_options: 
  chunk_output_type: inline
---

# Code Chunk 1: Load packages, prepare and inspect the data.
## Part A: Package loading, data import, and transformation.
```{r library}
# Import libraries
library(C50)
library(psych)
library(RWeka)
library(caret)
library(rminer)
library(matrixStats)
library(knitr)
library(arules)
library(tidyverse)

# Import dataset
mydir <- getwd()
setwd(mydir)

wmartvisits <- read.csv(file = "Walmart_visits_7trips.csv", stringsAsFactors = FALSE, fileEncoding = "latin1")

# Inspect the dataset
str(wmartvisits)

# Convert categorical variables to factors
wmartvisits <- wmartvisits %>%
  mutate(
    TripType = as.factor(TripType),
    DOW = as.factor(DOW)
  )

# Show summary of the mutated dataset
summary(wmartvisits)
```

## Part B: Understand this data set using correlation analysis.
```{r correlation-analysis}

```

## Part C: Build a descriptive C5.0 decision tree using the entire data set, Prune the tree so that the number of tree leaves is smaller than 15, Plot the tree and show summary of the model to view tree rules and confusion matrix.
```{r C5.0-decisiontree}

```