---
title: "Decision Tree, Naïve Bayes & Cross-validation"
author: "Jamison Orton"
date: "2026-06-16"
output: 
  html_document: 
    number_sections: true
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Task 1: Data and Decision Tree
## Question 1: Setup and starting code
### Part A - Set up, data import and inspection code
```{r library}
library(C50)
library(caret)
library(rminer)
library(tictoc)
library(tidyverse)
```
```{r setup-import}
mydir <- getwd()
setwd(mydir)

cd <- read.csv(file = "CD_additional_balanced-1.csv", stringsAsFactors = TRUE)
```
```{r structure-summary}
cd %>% str()

cd %>% summary()
```
### Part B - Data preparation
``` {r partition-data}
set.seed(123)

train_index <- createDataPartition(cd$y, p = 0.70, list = FALSE)

train <- cd[train_index, ]
test <- cd[-train_index, ]

train %>% summary()
test %>% summary()
```
### Part C - Target variable
``` {r target-variable-distributions}
cd %>% 
  count(y) %>% 
  mutate(percentage = round(n / sum(n) * 100, 2))

train %>% 
  count(y) %>% 
  mutate(percentage = round(n / sum(n) * 100, 2))

test %>% 
  count(y) %>% 
  mutate(percentage = round(n / sum(n) * 100, 2))
```

## Question 2: Simple Decision Tree Training and Testing

### Part A - 
``` {r train-base-model}
# Train default C5.0 model
c50_default <- C5.0(
  y ~ .,
  data = train
)

# Show model information
c50_default

# Show model summary
summary(c50_default)
```
```{r default-c50-evaluation}
# Predictions
train_pred_c50_default <- predict(c50_default, train)
test_pred_c50_default <- predict(c50_default, test)

# Confusion matrices
confusionMatrix(train_pred_c50_default, train$y)
confusionMatrix(test_pred_c50_default, test$y)

# Classification evaluation metrics
mmetric(
  y = train$y,
  x = train_pred_c50_default,
  metric = c("ACC", "PRECISION", "RECALL")
)

mmetric(
  y = test$y,
  x = test_pred_c50_default,
  metric = c("ACC", "PRECISION", "RECALL")
)
```

### Part B -
``` {r train-smaller-cf}
train_c50_model <- function(cf_value) {
  C5.0(
    y ~ .,
    data = train,
    control = C5.0Control(
      CF = cf_value,
      earlyStopping = FALSE,
      noGlobalPruning = FALSE
    )
  )
}

model_cf_001 <- train_c50_model(0.001)
```
```{r plot-smaller-model}
plot(model_cf_001)
```
```{r c50-reduced-evaluation}
# Generate predictions for training and testing sets
train_pred_cf_001 <- predict(model_cf_001, train)
test_pred_cf_001 <- predict(model_cf_001, test)

# Confusion matrices
confusionMatrix(train_pred_cf_001, train$y)
confusionMatrix(test_pred_cf_001, test$y)

# Classification evaluation metrics
mmetric(
  y = train$y,
  x = train_pred_cf_001,
  metric = c("ACC", "PRECISION", "RECALL")
)

mmetric(
  y = test$y,
  x = test_pred_cf_001,
  metric = c("ACC", "PRECISION", "RECALL")
)
```

## Question 3: Simple Naïve Bayes Model Training and Testing

### Part A -


### Part B - 


## Question 4: Create a Named Cross-validation Function – cv_function

### Part A - 


### Part B -


### Part C - 


### Part D - 


## Question 5: 5-fold and 10-fold C5.0 and naive Bayes evaluation performance with cv_function

### Part A -


# Task 2: Reflections
