---
title: 'A6: Blackbox Methods, KNN for Numeric Prediction tasks'
author: "Jamison Orton"
date: "2026-07-07"
output: 
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 15
    fig_height: 10
editor_options: 
  chunk_output_type: inline
---

# Code Chunk 1: Package load, data import, inspection and partitioning.
## Part A: Load required packages.
```{r library}
library(caret)
library(RWeka)
library(kernlab)
library(rminer)
library(matrixStats)
library(knitr)
library(tidyverse)
```
## Part B: Import NA_sales and partition dataset to the training and testing sets.
```{r import-partition}
mydir <- getwd()
setwd(mydir)

nasales <- read.csv(file = "NA_sales_filtered.csv", stringsAsFactors = FALSE, fileEncoding = "latin1")

# Remove the Name variable from the data frame
nasales_model <- nasales %>% 
  select(-Name)

nasales_model <- nasales_model %>% 
  mutate(across(where(is.character), as.factor))

df <- nasales_model
target <- "NA_Sales"
seedVal <- 500
metrics_list <- c("MAE","RMSE","MAPE","RMSPE","RAE","RRSE","R2")

# Create a 70/30 training/testing split based on the target variable
set.seed(seedVal)

train_index <- createDataPartition(
  nasales_model$NA_Sales, 
  p = 0.70, 
  list = FALSE
)

# Partition the data
train <- nasales_model[train_index, ]
test <- nasales_model[-train_index, ]

# Inspect the dataset
str(nasales)
summary(nasales_model)
glimpse(nasales_model)

# Check dimensions of full, training, and testing datasets
dim(nasales_model)
dim(train)
dim(test)
```

This section builds and evaluates two MultilayerPerceptron models: one using the default settings and one using two hidden layers with a changed learning rate.

# Code Chunk 2: Build and evaluate neural network models for numeric prediction tasks.
## Part A: Build and evaluate MLP models for numeric prediction with the video games sales data.
### Part 1: Build an MLP model on MultilayerPerceptron()’s default setting on the training set. Evaluate the model performance on the training set and testing set.
```{r mlp}
# Designate a shortened name MLP for the MultilayerPerceptron ANN method in RWeka
MLP <- make_Weka_classifier("weka/classifiers/functions/MultilayerPerceptron")

# Build MLP model on the training set using default settings
mlp_default <- MLP(NA_Sales ~ ., data = train)

# Make predictions on the training set
train_pred_default <- predict(mlp_default, newdata = train)

# Make predictions on the testing set
test_pred_default <- predict(mlp_default, newdata = test)

# Evaluate training set performance
train_performance_default <- mmetric(
  y = train$NA_Sales,
  x = train_pred_default,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_default <- mmetric(
  y = test$NA_Sales,
  x = test_pred_default,
  metric = metrics_list
)

# Combine performance results
mlp_default_results <- rbind(
  Training = train_performance_default,
  Testing = test_performance_default
)

# Display results
kable(mlp_default_results, digits = 2)
```

### Part 2: Build a two-hidden-layer MLP model and change one of the other hyper-parameter values.
```{r mlp-two-hidden-layer}
# Change learning rate from the default
l2 <- 0.1

# Build a two-hidden-layer MLP model
model_2 <- MLP(
  NA_Sales ~ .,
  data = train,
  control = Weka_control(
    L = l2,
    H = "5,5"
  )
)

# View model
model_2

# Make predictions on the training set
train_pred_2 <- predict(model_2, newdata = train)

# Make predictions on the testing set
test_pred_2 <- predict(model_2, newdata = test)

# Evaluate training set performance
train_performance_2 <- mmetric(
  y = train$NA_Sales,
  x = train_pred_2,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_2 <- mmetric(
  y = test$NA_Sales,
  x = test_pred_2,
  metric = metrics_list
)

# Combine performance results
model_2_performance <- rbind(
  Training = train_performance_2,
  Testing = test_performance_2
)

# Display results
kable(model_2_performance, digits = 4)
```

This section builds and evaluates three ksvm models: a default model, a model with a different kernel, and a model with a larger cost value.

# Code chunk 3: Build and evaluate SVM (ksvm) models for numeric prediction tasks.
## Part A: Build and evaluate ksvm models for numeric prediction with the video game sales data.
### Part 1: Build a model on ksvm()’s default setting on the training set. Evaluate the model performance on the training set and testing set.
```{r ksvm}
# Build ksvm model on the training set using default settings
ksvm_default <- ksvm(NA_Sales ~ ., data = train)

# Make predictions on the training set
train_pred_ksvm_default <- predict(ksvm_default, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_default <- predict(ksvm_default, newdata = test)

# Evaluate training set performance
train_performance_ksvm_default <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_default,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_default <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_default,
  metric = metrics_list
)

# Combine performance results
ksvm_default_results <- rbind(
  Training = train_performance_ksvm_default,
  Testing = test_performance_ksvm_default
)

# Display results
kable(ksvm_default_results, digits = 2)
```

### Part 2: Build a ksvm model using a different kernel function on the training set. Use default C value.
```{r ksvm-different-kernel}
# Build ksvm model using a different kernel function
# vanilladot is the linear kernel
ksvm_linear <- ksvm(
  NA_Sales ~ .,
  data = train,
  kernel = "vanilladot"
)

# View model
ksvm_linear

# Make predictions on the training set
train_pred_ksvm_linear <- predict(ksvm_linear, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_linear <- predict(ksvm_linear, newdata = test)

# Evaluate training set performance
train_performance_ksvm_linear <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_linear,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_linear <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_linear,
  metric = metrics_list
)

# Combine performance results
ksvm_linear_results <- rbind(
  Training = train_performance_ksvm_linear,
  Testing = test_performance_ksvm_linear
)

# Display results
kable(ksvm_linear_results, digits = 2)
```

### Part 3: Build a ksvm model using a different cost value where C > 1.
```{r ksvm-different-cost}
# Set cost value greater than 1
c_value <- 5

# Build ksvm model using a different cost value
ksvm_cost <- ksvm(
  NA_Sales ~ .,
  data = train,
  C = c_value
)

# View model
ksvm_cost

# Make predictions on the training set
train_pred_ksvm_cost <- predict(ksvm_cost, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_cost <- predict(ksvm_cost, newdata = test)

# Evaluate training set performance
train_performance_ksvm_cost <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_cost,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_cost <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_cost,
  metric = metrics_list
)

# Combine performance results
ksvm_cost_results <- rbind(
  Training = train_performance_ksvm_cost,
  Testing = test_performance_ksvm_cost
)

# Display results
kable(ksvm_cost_results, digits = 2)
```

This section builds and evaluates several IBk KNN models using default settings, a different K value, weighted voting, and automatic K selection.

# Code chunk 4: Build and evaluate IBk models for numeric prediction tasks
## Part A: Build and evaluate IBk models for numeric prediction with the video game sales data
### Part 1: Build a model on IBk()'s default setting on the training set. Evaluate the model performance on the training set and testing set.

```{r ibk-default}
# Designate a shortened name IBk for the IBk KNN method in RWeka
IBk <- make_Weka_classifier("weka/classifiers/lazy/IBk")

# Build IBk model on the training set using default settings
ibk_default <- IBk(
  NA_Sales ~ .,
  data = train
)

# View model
ibk_default

# Make predictions on the training set
train_pred_ibk_default <- predict(ibk_default, newdata = train)

# Make predictions on the testing set
test_pred_ibk_default <- predict(ibk_default, newdata = test)

# Evaluate training set performance
train_performance_ibk_default <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ibk_default,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ibk_default <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ibk_default,
  metric = metrics_list
)

# Combine performance results
ibk_default_results <- rbind(
  Training = train_performance_ibk_default,
  Testing = test_performance_ibk_default
)

# Display results
kable(ibk_default_results, digits = 2)
```

### Part 2: Build an IBk model using a different K value
```{r ibk-different-k}
# Set K value
k_value <- 5

# Build IBk model using a different K value
ibk_k <- IBk(
  NA_Sales ~ .,
  data = train,
  control = Weka_control(K = k_value)
)

# View model
ibk_k

# Make predictions on the training set
train_pred_ibk_k <- predict(ibk_k, newdata = train)

# Make predictions on the testing set
test_pred_ibk_k <- predict(ibk_k, newdata = test)

# Evaluate training set performance
train_performance_ibk_k <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ibk_k,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ibk_k <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ibk_k,
  metric = metrics_list
)

# Combine performance results
ibk_k_results <- rbind(
  Training = train_performance_ibk_k,
  Testing = test_performance_ibk_k
)

# Display results
kable(ibk_k_results, digits = 2)
```

### Part 3: Build an IBk model using a weighted voting approach
```{r ibk-weighted-voting}
# Build IBk model using weighted voting
ibk_weighted <- IBk(
  NA_Sales ~ .,
  data = train,
  control = Weka_control(I = TRUE)
)

# View model
ibk_weighted

# Make predictions on the training set
train_pred_ibk_weighted <- predict(ibk_weighted, newdata = train)

# Make predictions on the testing set
test_pred_ibk_weighted <- predict(ibk_weighted, newdata = test)

# Evaluate training set performance
train_performance_ibk_weighted <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ibk_weighted,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ibk_weighted <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ibk_weighted,
  metric = metrics_list
)

# Combine performance results
ibk_weighted_results <- rbind(
  Training = train_performance_ibk_weighted,
  Testing = test_performance_ibk_weighted
)

# Display results
kable(ibk_weighted_results, digits = 2)
```

### Part 4: Build an IBk model by automatically selecting K
```{r ibk-auto-k}
# Build IBk model using automatic K selection
# K sets the maximum K to consider, and X = TRUE enables cross-validation to choose K
ibk_auto_k <- IBk(
  NA_Sales ~ .,
  data = train,
  control = Weka_control(
    K = 10,
    X = TRUE
  )
)

# View model
ibk_auto_k

# Make predictions on the training set
train_pred_ibk_auto_k <- predict(ibk_auto_k, newdata = train)

# Make predictions on the testing set
test_pred_ibk_auto_k <- predict(ibk_auto_k, newdata = test)

# Evaluate training set performance
train_performance_ibk_auto_k <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ibk_auto_k,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ibk_auto_k <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ibk_auto_k,
  metric = metrics_list
)

# Combine performance results
ibk_auto_k_results <- rbind(
  Training = train_performance_ibk_auto_k,
  Testing = test_performance_ibk_auto_k
)

# Display results
kable(ibk_auto_k_results, digits = 2)
```

This section defines one general cross-validation function that can be reused with different model methods.

# Code chunk 5: Cross-validation function for numeric prediction models
## Part A/B: Define a named CV function that returns fold-by-fold metrics, means, and standard deviations
```{r cv-function}
cv_function <- function(df, target, nFolds, seedVal, method, metrics_list)
{
  # Create folds using the assigned seed value
  set.seed(seedVal)
  folds <- createFolds(df[[target]], k = nFolds)

  # Create formula from target variable
  model_formula <- as.formula(paste(target, "~ ."))

  # Cross-validation loop
  cv_results <- lapply(folds, function(x)
  {
    # Split into training and validation folds
    train_fold <- df[-x, ]
    test_fold <- df[x, ]

    # Build model using the method passed into the function
    pred_model <- method(
      model_formula,
      data = train_fold
    )

    # Predict on validation fold
    pred <- predict(pred_model, newdata = test_fold)

    # Evaluate performance
    mmetric(
      y = test_fold[[target]],
      x = pred,
      metric = metrics_list
    )
  })

  # Convert fold results into a matrix
  cv_results_m <- as.matrix(as.data.frame(cv_results))

  # Rename columns as Fold 1, Fold 2, etc.
  colnames(cv_results_m) <- paste0("Fold_", 1:nFolds)

  # Calculate mean and standard deviation across folds
  cv_mean <- rowMeans(cv_results_m)
  cv_sd <- rowSds(cv_results_m)

  # Combine fold-by-fold results with mean and standard deviation
  cv_all <- cbind(
    cv_results_m,
    Mean = cv_mean,
    Sd = cv_sd
  )

  # Display full CV table
  kable(cv_all, digits = 2)
}
```

This section uses 3-fold cross-validation to evaluate the default settings of three numeric prediction models: MultilayerPerceptron, ksvm, and IBk. The same cross-validation function is used for each model so that the fold-by-fold performance metrics, means, and standard deviations can be compared across models.

# Code Chunk 6: 3 fold cross-validation of MLP, ksvm and IBk models 
```{r three-fold-cv-default-models}
# Make sure the Weka classifiers are defined
MLP <- make_Weka_classifier("weka/classifiers/functions/MultilayerPerceptron")
IBk <- make_Weka_classifier("weka/classifiers/lazy/IBk")

# 3-fold CV for default MultilayerPerceptron
cv_function(
  df = df,
  target = target,
  nFolds = 3,
  seedVal = seedVal,
  method = MLP,
  metrics_list = metrics_list
)

# 3-fold CV for default ksvm
cv_function(
  df = df,
  target = target,
  nFolds = 3,
  seedVal = seedVal,
  method = ksvm,
  metrics_list = metrics_list
)

# 3-fold CV for default IBk
cv_function(
  df = df,
  target = target,
  nFolds = 3,
  seedVal = seedVal,
  method = IBk,
  metrics_list = metrics_list
)
```

Task II Reflections

1. KSVM model performance results

For the ksvm models, changing the parameters did affect the performance. The default ksvm model performed pretty well overall. When I changed the kernel to the linear kernel, the performance got worse because the RMSE increased and the R² decreased. This tells me that the relationship between the variables and NA_Sales is probably not very linear. A nonlinear kernel seems to fit this data better.

When I changed the cost value to C = 5, the model improved compared to the default ksvm model. The RMSE decreased and the R² increased slightly. This means the model was able to fit the data better without hurting the testing performance. I think the higher cost value helped because it penalized errors more, so the model tried harder to fit the training data. In this case, that helped rather than causing serious overfitting. Overall, the ksvm model with C = 5 seemed to be the strongest ksvm model.

2. MLP model performance results

For the MLP models, the two-hidden-layer model performed better than the default model. The default MLP had higher testing error and a lower R². After changing the model to have two hidden layers and lowering the learning rate, the RMSE decreased and the R² increased. This means the adjusted MLP model made better predictions on the testing data.

I think this happened because the two-hidden-layer model was able to capture more complex patterns in the data. The lower learning rate may have also helped the model learn more carefully instead of changing the weights too quickly. The tradeoff is that the adjusted MLP probably took longer to run. I did not record exact runtimes, but a more complex neural network usually takes more time because it has more calculations to complete. In this case, the slower speed seems acceptable because the model performance improved.

3. Overall interpretation of the models

This assignment helped me see how much model performance can change when hyperparameters are adjusted. The ksvm models showed that the default nonlinear approach worked better than the linear kernel, which suggests that the sales data has nonlinear relationships. The ksvm model with the higher cost value had the best performance among the ksvm models, so I would probably recommend that version if I were explaining the results to a supervisor.

The MLP models also showed that changing the model structure can help. The two-hidden-layer MLP did better than the default MLP, which suggests that the neural network benefited from having more flexibility. However, I would also mention that this model may take longer to train and evaluate.

The IBk models showed the clearest example of overfitting. The default K = 1 model did extremely well on the training data but much worse on the testing data. This means it mostly memorized the training data instead of learning a pattern that worked well on new data. Increasing or automatically selecting K helped make the model more reasonable, but I would still be cautious about relying on IBk compared with the better ksvm and MLP models. Overall, I learned that testing performance matters more than training performance because a model that looks perfect on training data may not actually generalize well.