---
title: 'A6: Blackbox Methods, KNN for Numeric Prediction tasks'
author: "Jamison Orton"
date: "2026-07-07"
output: 
  html_document:
    number_sections: yes
    toc: yes
    fig_width: 15
    fig_height: 10
editor_options: 
  chunk_output_type: inline
---

# Code Chunk 1: Package load, data import, inspection and partitioning.
## Part A: Load required packages.
```{r library}
library(caret)
library(RWeka)
library(kernlab)
library(rminer)
library(matrixStats)
library(knitr)
library(tidyverse)
```
## Part B: Import NA_sales and partition dataset to the training and testing sets.
```{r import-partition}
mydir <- getwd()
setwd(mydir)

nasales <- read.csv(file = "NA_sales_filtered.csv", stringsAsFactors = FALSE, fileEncoding = "latin1")

# Remove the Name variable from the data frame
nasales_model <- nasales %>% 
  select(-Name)

nasales_model <- nasales_model %>% 
  mutate(across(where(is.character), as.factor))

df <- nasales_model
target <- "NA_Sales"
seedVal <- 500
metrics_list <- c("MAE","RMSE","MAPE","RMSPE","RAE","RRSE","R2")

# Create a 70/30 training/testing split based on the target variable
set.seed(seedVal)

train_index <- createDataPartition(
  nasales_model$NA_Sales, 
  p = 0.70, 
  list = FALSE
)

# Partition the data
train <- nasales_model[train_index, ]
test <- nasales_model[-train_index, ]
```

# Code Chunk 2: Build and evaluate neural network models for numeric prediction tasks.
## Part A: Build and evaluate MLP models for numeric prediction with the video games sales data.
### Part 1: Build an MLP model on MultilayerPerceptron()’s default setting on the training set. Evaluate the model performance on the training set and testing set.
```{r mlp}
# Designate a shortened name MLP for the MultilayerPerceptron ANN method in RWeka
MLP <- make_Weka_classifier("weka/classifiers/functions/MultilayerPerceptron")

# Build MLP model on the training set using default settings
mlp_default <- MLP(NA_Sales ~ ., data = train)

# Make predictions on the training set
train_pred_default <- predict(mlp_default, newdata = train)

# Make predictions on the testing set
test_pred_default <- predict(mlp_default, newdata = test)

# Evaluate training set performance
train_performance_default <- mmetric(
  y = train$NA_Sales,
  x = train_pred_default,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_default <- mmetric(
  y = test$NA_Sales,
  x = test_pred_default,
  metric = metrics_list
)

# Combine performance results
mlp_default_results <- rbind(
  Training = train_performance_default,
  Testing = test_performance_default
)

# Display results
kable(mlp_default_results, digits = 2)
```

### Part 2: Build a two-hidden-layer MLP model and change one of the other hyper-parameter values.
```{r mlp-two-hidden-layer}
# Change learning rate from the default
l2 <- 0.1

# Build a two-hidden-layer MLP model
model_2 <- MLP(
  NA_Sales ~ .,
  data = train,
  control = Weka_control(
    L = l2,
    H = "1,1"
  )
)

# View model
model_2

# Make predictions on the training set
train_pred_2 <- predict(model_2, newdata = train)

# Make predictions on the testing set
test_pred_2 <- predict(model_2, newdata = test)

# Evaluate training set performance
train_performance_2 <- mmetric(
  y = train$NA_Sales,
  x = train_pred_2,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_2 <- mmetric(
  y = test$NA_Sales,
  x = test_pred_2,
  metric = metrics_list
)

# Combine performance results
model_2_performance <- rbind(
  Training = train_performance_2,
  Testing = test_performance_2
)

# Display results
kable(model_2_performance, digits = 4)
```

# Code chunk 3: Build and evaluate SVM (ksvm) models for numeric prediction tasks.
## Part A: Build and evaluate ksvm models for numeric prediction with the video game sales data.
### Part 1: Build a model on ksvm()’s default setting on the training set. Evaluate the model performance on the training set and testing set.
```{r ksvm}
# Build ksvm model on the training set using default settings
ksvm_default <- ksvm(NA_Sales ~ ., data = train)

# Make predictions on the training set
train_pred_ksvm_default <- predict(ksvm_default, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_default <- predict(ksvm_default, newdata = test)

# Evaluate training set performance
train_performance_ksvm_default <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_default,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_default <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_default,
  metric = metrics_list
)

# Combine performance results
ksvm_default_results <- rbind(
  Training = train_performance_ksvm_default,
  Testing = test_performance_ksvm_default
)

# Display results
kable(ksvm_default_results, digits = 2)
```

### Part 2: Build a ksvm model using a different kernel function on the training set. Use default C value.
```{r ksvm-different-kernel}
# Build ksvm model using a different kernel function
# vanilladot is the linear kernel
ksvm_linear <- ksvm(
  NA_Sales ~ .,
  data = train,
  kernel = "vanilladot"
)

# View model
ksvm_linear

# Make predictions on the training set
train_pred_ksvm_linear <- predict(ksvm_linear, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_linear <- predict(ksvm_linear, newdata = test)

# Evaluate training set performance
train_performance_ksvm_linear <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_linear,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_linear <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_linear,
  metric = metrics_list
)

# Combine performance results
ksvm_linear_results <- rbind(
  Training = train_performance_ksvm_linear,
  Testing = test_performance_ksvm_linear
)

# Display results
kable(ksvm_linear_results, digits = 2)
```

### Part 3: Build a ksvm model using a different cost value where C > 1.
```{r ksvm-different-cost}
# Set cost value greater than 1
c_value <- 5

# Build ksvm model using a different cost value
ksvm_cost <- ksvm(
  NA_Sales ~ .,
  data = train,
  C = c_value
)

# View model
ksvm_cost

# Make predictions on the training set
train_pred_ksvm_cost <- predict(ksvm_cost, newdata = train)

# Make predictions on the testing set
test_pred_ksvm_cost <- predict(ksvm_cost, newdata = test)

# Evaluate training set performance
train_performance_ksvm_cost <- mmetric(
  y = train$NA_Sales,
  x = train_pred_ksvm_cost,
  metric = metrics_list
)

# Evaluate testing set performance
test_performance_ksvm_cost <- mmetric(
  y = test$NA_Sales,
  x = test_pred_ksvm_cost,
  metric = metrics_list
)

# Combine performance results
ksvm_cost_results <- rbind(
  Training = train_performance_ksvm_cost,
  Testing = test_performance_ksvm_cost
)

# Display results
kable(ksvm_cost_results, digits = 2)
```

# Code chunk 4: Build and evaluate KNN (IBk) models for numeric prediction tasks.
## Part A: Build and evaluate IBk models for numeric prediction with the video game sales data.
### Part 1: Build a model on IBk()’s default setting on the training set.