---
title: "Decision Tree Classification and Evaluation"
author: "Jamison Orton"
date: "2026-06-09"
output: 
  html_document: 
    number_sections: true
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Task 1: Data and Decision Tree

## Part 1 - Set up, data import and inspection code
```{r library}
library(C50)
library(caret)
library(rminer)
library(tictoc)
library(tidyverse)
```
```{r Set up, data import and inspection}
cd <- read.csv(file = "CD_additional_balanced-1.csv", stringsAsFactors = FALSE)
```
```{r structure aka str}
cd %>% str()
```
```{r transform char to factor and show overall structure and summary}
cd <- cd %>% mutate(across(where(is.character), as.factor))

cd %>% str()

cd %>% summary()
```
## Part 2 - Target variable
``` {r show count for target variable and percentage of instances belonging to that level}
cd %>% count(y) %>% mutate(percentage = round(n / sum(n) * 100, 2))
```

## Part 3 - Data preparation
``` {r Partition data set for simple hold-out classification model building and evaluation (show summary of train and test sets)}
set.seed(123)

train_index <- createDataPartition(cd$y, p = 0.70, list = FALSE)

train <- cd[train_index, ]
test <- cd[-train_index, ]

train %>% summary()
test %>% summary()
```
``` {r show count and distributions of y in the train set and in the test set}
train %>%
  count(y) %>%
  mutate(percent = round(n / sum(n) * 100, 2))

test %>%
  count(y) %>%
  mutate(percent = round(n / sum(n) * 100, 2))
```
## Part 4 - Train Decision Trees to classify y
``` {r train 7 models at various CF}
train_c50_model <- function(cf_value) {
  C5.0(
    y ~ .,
    data = train,
    control = C5.0Control(
      CF = cf_value,
      earlyStopping = FALSE,
      noGlobalPruning = FALSE
    )
  )
}

model_cf_97 <- train_c50_model(0.97)
model_cf_35 <- train_c50_model(0.35)
model_cf_12 <- train_c50_model(0.12)
model_cf_08 <- train_c50_model(0.08)
model_cf_05 <- train_c50_model(0.05)
model_cf_025 <- train_c50_model(0.025)
model_cf_001 <- train_c50_model(0.001)
```

## Part 5 - Model Information
### Part A
``` {r part a}
tree_sizes <- data.frame(
  model = c(
    "model_cf_97",
    "model_cf_35",
    "model_cf_12",
    "model_cf_08",
    "model_cf_05",
    "model_cf_025",
    "model_cf_001"
  ),
  
  CF = c(0.97, 0.35, 0.12, 0.08, 0.05, 0.025, 0.001),
  
  tree_size = c(
    model_cf_97$size,
    model_cf_35$size,
    model_cf_12$size,
    model_cf_08$size,
    model_cf_05$size,
    model_cf_025$size,
    model_cf_001$size
  )
)

tree_sizes
```
### Part B
Tree complexity is based on tree size. Tree size represents the number of terminal decision rules in the model. A tree with more leaf nodes is more complex reuiring more specific rules for classifying observations. A tree with fewer leaf nodes is less complex and uses fewer rules and is easier to interpret.

### Part C
```{r plot-least-complex-tree, fig.height=8, fig.width=20}
least_complex_model_name <- tree_sizes %>%
  arrange(tree_size) %>%
  slice(1) %>%
  pull(model)
least_complex_model <- get(least_complex_model_name)

least_complex_model_name
plot(least_complex_model)
```
### Part D
``` {r show least_complex_model and then explain steps}
least_complex_model
```
For the instance with nr.employed = 6000 and duration = 500, I used the least complex decision tree to follow the model's decision rules. The least complex tree begins by checking the value of nr.employed if it is <= 5076.2 then it will go to duration. In the instance of nr.employed = 6000 and duration = 500 the model predicts yes as it is yes 80% of the time.

## Part 6 - Predict on the Train and Test sets with each trained model
``` {r generate-train-test-predictions}
# Predictions on the Train set
pred_train_cf_97 <- predict(model_cf_97, newdata = train)
pred_train_cf_35 <- predict(model_cf_35, newdata = train)
pred_train_cf_12 <- predict(model_cf_12, newdata = train)
pred_train_cf_08 <- predict(model_cf_08, newdata = train)
pred_train_cf_05 <- predict(model_cf_05, newdata = train)
pred_train_cf_025 <- predict(model_cf_025, newdata = train)
pred_train_cf_001 <- predict(model_cf_001, newdata = train)

# Predictions on the Test set
pred_test_cf_97 <- predict(model_cf_97, newdata = test)
pred_test_cf_35 <- predict(model_cf_35, newdata = test)
pred_test_cf_12 <- predict(model_cf_12, newdata = test)
pred_test_cf_08 <- predict(model_cf_08, newdata = test)
pred_test_cf_05 <- predict(model_cf_05, newdata = test)
pred_test_cf_025 <- predict(model_cf_025, newdata = test)
pred_test_cf_001 <- predict(model_cf_001, newdata = test)
```

## Part 7 - Generate Confusion Matrices for each trained Model for both Train and Test
``` {r confusion-matrices-train-test}
# Confusion matrices for the Train set
cm_train_cf_97 <- confusionMatrix(pred_train_cf_97, train$y)
cm_train_cf_35 <- confusionMatrix(pred_train_cf_35, train$y)
cm_train_cf_12 <- confusionMatrix(pred_train_cf_12, train$y)
cm_train_cf_08 <- confusionMatrix(pred_train_cf_08, train$y)
cm_train_cf_05 <- confusionMatrix(pred_train_cf_05, train$y)
cm_train_cf_025 <- confusionMatrix(pred_train_cf_025, train$y)
cm_train_cf_001 <- confusionMatrix(pred_train_cf_001, train$y)

# Confusion matrices for the Test set
cm_test_cf_97 <- confusionMatrix(pred_test_cf_97, test$y)
cm_test_cf_35 <- confusionMatrix(pred_test_cf_35, test$y)
cm_test_cf_12 <- confusionMatrix(pred_test_cf_12, test$y)
cm_test_cf_08 <- confusionMatrix(pred_test_cf_08, test$y)
cm_test_cf_05 <- confusionMatrix(pred_test_cf_05, test$y)
cm_test_cf_025 <- confusionMatrix(pred_test_cf_025, test$y)
cm_test_cf_001 <- confusionMatrix(pred_test_cf_001, test$y)
```
```{r show-confusion-matrices}
cm_train_cf_97
cm_train_cf_35
cm_train_cf_12
cm_train_cf_08
cm_train_cf_05
cm_train_cf_025
cm_train_cf_001
cm_test_cf_97
cm_test_cf_35
cm_test_cf_12
cm_test_cf_08
cm_test_cf_05
cm_test_cf_025
cm_test_cf_001
```

## Part 8 - Generate Evaluation Metrics for each Model
```{r evaluation-metrics}
# Metrics requested: F1, Accuracy, Precision, and Recall/TPR
metric_list <- c("F1", "ACC", "PRECISION", "TPR")

# Train set metrics
metrics_train_cf_97 <- mmetric(train$y, pred_train_cf_97, metric = metric_list)
metrics_train_cf_35 <- mmetric(train$y, pred_train_cf_35, metric = metric_list)
metrics_train_cf_12 <- mmetric(train$y, pred_train_cf_12, metric = metric_list)
metrics_train_cf_08 <- mmetric(train$y, pred_train_cf_08, metric = metric_list)
metrics_train_cf_05 <- mmetric(train$y, pred_train_cf_05, metric = metric_list)
metrics_train_cf_025 <- mmetric(train$y, pred_train_cf_025, metric = metric_list)
metrics_train_cf_001 <- mmetric(train$y, pred_train_cf_001, metric = metric_list)

# Test set metrics
metrics_test_cf_97 <- mmetric(test$y, pred_test_cf_97, metric = metric_list)
metrics_test_cf_35 <- mmetric(test$y, pred_test_cf_35, metric = metric_list)
metrics_test_cf_12 <- mmetric(test$y, pred_test_cf_12, metric = metric_list)
metrics_test_cf_08 <- mmetric(test$y, pred_test_cf_08, metric = metric_list)
metrics_test_cf_05 <- mmetric(test$y, pred_test_cf_05, metric = metric_list)
metrics_test_cf_025 <- mmetric(test$y, pred_test_cf_025, metric = metric_list)
metrics_test_cf_001 <- mmetric(test$y, pred_test_cf_001, metric = metric_list)
```
```{r evaluation-metrics-dataframe}
evaluation_results <- data.frame(
  dataset = c(
    rep("Train", 7),
    rep("Test", 7)
  ),

  model = rep(
    c(
      "model_cf_97",
      "model_cf_35",
      "model_cf_12",
      "model_cf_08",
      "model_cf_05",
      "model_cf_025",
      "model_cf_001"
    ),
    2
  ),

  CF = rep(
    c(0.97, 0.35, 0.12, 0.08, 0.05, 0.025, 0.001),
    2
  ),

  F1 = c(
    metrics_train_cf_97["F1"],
    metrics_train_cf_35["F1"],
    metrics_train_cf_12["F1"],
    metrics_train_cf_08["F1"],
    metrics_train_cf_05["F1"],
    metrics_train_cf_025["F1"],
    metrics_train_cf_001["F1"],
    metrics_test_cf_97["F1"],
    metrics_test_cf_35["F1"],
    metrics_test_cf_12["F1"],
    metrics_test_cf_08["F1"],
    metrics_test_cf_05["F1"],
    metrics_test_cf_025["F1"],
    metrics_test_cf_001["F1"]
  ),

  Accuracy = c(
    metrics_train_cf_97["ACC"],
    metrics_train_cf_35["ACC"],
    metrics_train_cf_12["ACC"],
    metrics_train_cf_08["ACC"],
    metrics_train_cf_05["ACC"],
    metrics_train_cf_025["ACC"],
    metrics_train_cf_001["ACC"],
    metrics_test_cf_97["ACC"],
    metrics_test_cf_35["ACC"],
    metrics_test_cf_12["ACC"],
    metrics_test_cf_08["ACC"],
    metrics_test_cf_05["ACC"],
    metrics_test_cf_025["ACC"],
    metrics_test_cf_001["ACC"]
  ),

  Precision = c(
    metrics_train_cf_97["PRECISION"],
    metrics_train_cf_35["PRECISION"],
    metrics_train_cf_12["PRECISION"],
    metrics_train_cf_08["PRECISION"],
    metrics_train_cf_05["PRECISION"],
    metrics_train_cf_025["PRECISION"],
    metrics_train_cf_001["PRECISION"],
    metrics_test_cf_97["PRECISION"],
    metrics_test_cf_35["PRECISION"],
    metrics_test_cf_12["PRECISION"],
    metrics_test_cf_08["PRECISION"],
    metrics_test_cf_05["PRECISION"],
    metrics_test_cf_025["PRECISION"],
    metrics_test_cf_001["PRECISION"]
  ),

  Recall = c(
    metrics_train_cf_97["TPR"],
    metrics_train_cf_35["TPR"],
    metrics_train_cf_12["TPR"],
    metrics_train_cf_08["TPR"],
    metrics_train_cf_05["TPR"],
    metrics_train_cf_025["TPR"],
    metrics_train_cf_001["TPR"],
    metrics_test_cf_97["TPR"],
    metrics_test_cf_35["TPR"],
    metrics_test_cf_12["TPR"],
    metrics_test_cf_08["TPR"],
    metrics_test_cf_05["TPR"],
    metrics_test_cf_025["TPR"],
    metrics_test_cf_001["TPR"]
  )
)

evaluation_results
```

## Part 9 - Show the feature importance for each Model
```{r feature-importance}
# Feature importance for each trained C5.0 decision tree
importance_cf_97 <- C5imp(model_cf_97)
importance_cf_35 <- C5imp(model_cf_35)
importance_cf_12 <- C5imp(model_cf_12)
importance_cf_08 <- C5imp(model_cf_08)
importance_cf_05 <- C5imp(model_cf_05)
importance_cf_025 <- C5imp(model_cf_025)
importance_cf_001 <- C5imp(model_cf_001)
importance_cf_97
importance_cf_35
importance_cf_12
importance_cf_08
importance_cf_05
importance_cf_025
importance_cf_001
```

## Part 10 (OPTIONAL) - Generate a Graph showing Train and Test Accuracy for each model
``` {r}

```

# Task 2: Reflections
Changing the CF hyperparameter affected the complexity of the decision tree models. In general, higher CF values allowed for more splits, creating larger and more complex trees. Lower CF values increased pruning creating smaller trees. In my results, the most complex model was the model with a CF of .97 creating 339 leaf nodes. The least complex was the model with a CF of .001 creating 13 leaf nodes. I defined complexity using the number of leaf nodes because each leaf node represents a final decision rule. Higher leaf nodes are harder to interpret, while a model with fewer nodes is easier to explain.

The best performing model on the training set was the model with CF of .97. This model had 339 leaf nodes and achieved a training accuracy of 94.38%. The best-performing model on the testing set was the model with CF of .12. The model had 59 leaf nodes and achieved a test accuracy of 88.5%.

Overall the relationship between complexity and performance showed that more complex trees tended to perform better on the training data but they did not always perform better on the testing data. This suggest that some of the larger trees may have overfit the training data. 

If I were presenting these results in a meeting, I would choose the model with CF of .05. While it does not have the highest training accuracy, it performs well on the testing set and has a manageable number of leaf nodes. This makes it easier to explain how the model makes predections.

The relationship between complexity and runtime is that more complex trees generally require more time to generate because the algorithm evaluates more possible splits. High runtime can be a problem for larger datasets because it slows model development and increases computing costs.