---
title: "Decision Tree Classification and Evaluation"
author: "Jamison Orton"
date: "2026-06-09"
output: 
  html_document: 
    number_sections: true
    toc: true
editor_options: 
  chunk_output_type: inline
---

# Task 1: Data and Decision Tree

## Part 1 - Set up, data import and inspection code
```{r library}
library(C50)
library(caret)
library(rminer)
library(tictoc)
library(tidyverse)
```
```{r Set up, data import and inspection}
cd <- read.csv(file = "CD_additional_balanced-1.csv", stringsAsFactors = FALSE)
```
```{r structure aka str}
cd %>% str()
```
```{r transform char to factor and show overall structure and summary}
cd <- cd %>% mutate(across(where(is.character), as.factor))

cd %>% str()

cd %>% summary()
```
## Part 2 - Target variable
``` {r show count for target variable and percentage of instances belonging to that level}
cd %>% count(y) %>% mutate(percentage = round(n / sum(n) * 100, 2))
```

## Part 3 - Data preparation
``` {r Partition data set for simple hold-out classification model building and evaluation (show summary of train and test sets)}
set.seed(123)

train_index <- createDataPartition(cd$y, p = 0.70, list = FALSE)

train <- cd[train_index, ]
test <- cd[-train_index, ]

train %>% summary()
test %>% summary()
```
``` {r show count and distributions of y in the train set and in the test set}
train %>%
  count(y) %>%
  mutate(percent = round(n / sum(n) * 100, 2))

test %>%
  count(y) %>%
  mutate(percent = round(n / sum(n) * 100, 2))
```
## Part 4 - Train Decision Trees to classify y
``` {r train 7 models at various CF}
c50_control <- C5.0Control(
  earlyStopping = FALSE,
  noGlobalPruning = FALSE
)

model_cf_97 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.97)
model_cf_35 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.35)
model_cf_12 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.12)
model_cf_08 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.08)
model_cf_05 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.05)
model_cf_025 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.025)
model_cf_001 <- C5.0(formula = y ~ ., data = train, control = c50_control, CF = 0.001)

model_cf_001
```

## Part 5 - Model Information
``` {r}

```

## Part 6 - Predict on the Train and Test sets with each trained model
``` {r}

```

## Part 7 - Generate Confusion Matrices for each trained Model for both Train and Test
``` {r}

```

## Part 8 - Generate Evaluation Metrics for each Model
``` {r}

```

## Part 9 - Show the feature importance for each Model
``` {r}

```

## Part 10 (OPTIONAL) - Generate a Graph showing Train and Test Accuracy for each model
``` {r}

```

# Task 2: Reflections