---
title: "Regression models, model fit, and prediction errors"
author: "Jamison Orton"
date: "2026-06-23"
output: 
  html_document: 
    number_sections: true
    toc: true
editor_options: 
  chunk_output_type: inline
---
# Task 1: Data and Regression Model
## Question 1: Setup and starting code
### Part A - Set up, data import and inspection code
```{r library}
library(psych)
library(rpart)
library(RWeka)
library(caret)
library(rminer)
library(matrixStats)
library(knitr)
library(tidyverse)
```
```{r setup-import}
mydir <- getwd()
setwd(mydir)

vgsales <- read.csv(file = "NA_sales_filtered.csv", stringsAsFactors = FALSE, fileEncoding = "latin1")

vgsales %>% str()

vgsales <- vgsales %>% 
  mutate(across(where(is.character) & !Name, as.factor))

vgsales %>% summary()
```

### Part B - Pairs panel distribution and correlation
``` {r pairs-panel}

```

### Part C - Remove Name Variable, Regression model summary
``` {r regression-model}

```

### Part D - Simple hold-out evaluation
``` {r hold-out}
set.seed(123)

vg_index <- createDataPartition(cd$y, p = 0.70, list = FALSE)

train <- cd[train_index, ]
test <- cd[-train_index, ]

train %>% summary()
test %>% summary()
```

### Part E - Structure and summaries
``` {r summaries}

```

## Question 2: lm, rpart and M5P model training and testing

## Question 3: Cross-validation of lm, rpart, and M5P NA_Sales prediction models

## Question 4: Improve the models by adding a quadratic term of Critic_Score

## Question 5: Improve the models with the log term of User_Count