---
title: "Flight Delay Case"
author: "Jamison Orton"
date: "October 2, 2026"
format:
  html:
    toc: true
---
## R Packages
```{r libraries}
library(tidyverse)
options(pillar.sigfig = 5)
```

## Q1
### Compute the mean, median, 90th percentile, and standard deviation of arrival delay minutes for RegionEx flights. Do the same for MDA flights.
```{r, math_stuff}
flightdf <- read.csv("flight_delay_clean.csv")

flightdf %>%
  filter(airline %in% c("RegionEx", "MDA")) %>%
  group_by(airline) %>%
  summarise(
    n = n(),
    mean_delay = round(mean(delay), 2),
    median_delay = round(median(delay), 2),
    perc_90 = round(as.numeric(quantile(delay, probs = 0.90)), 2),
    sd_delay = round(sd(delay), 2)
  )
```
### Contractual obligations aside, which measure of central tendency would be most appropriate for comparing airline performance?

Removing contractual obligation, using the median would most likely be the best measure of central tendency for comparing typical airline performance. RegionEx has a mean delay of 15.66 minutes but a median delay of only 9. This difference coupled with the high standard deviation of 27.65 shows that there are several large outliers skyrocketing the mean higher but that is not the "typical" flight experience. However, if the goal is to measure total passenger impact, the mean would be a more appropriate comparison because it reflects the influence of extreme delays.

## Q2
### Inspect the distribution of RegionEx’s arrival delays by constructing a histogram of the number of arrival delay minutes of RegionEx’s flights. Do the same for MDA’s flights. Hint: use facet_wrap().
```{r}
flightdf %>%
  filter(airline %in% c("RegionEx", "MDA")) %>%
  ggplot(aes(x = delay)) +
  geom_histogram(
    aes(fill = airline),
    binwidth = 20,
    color = "black"
  ) +
  facet_wrap(~airline) +
  labs(
    title = "Arrival Delay Distributions by Airline",
    x = "Arrival delay minutes",
    y = "Number of flights"
  ) +
  theme_minimal()
```

### How do these two distributions compare?
Both of these histograms are very heavily right skewed. This shows 

## Q3
So far you have considered airline performance in terms of minutes delayed. However, the performance metrics, as noted in the case description, also include the percentage of delayed flights. Let’s verify that MDA’s COO is correct: does RegionEx have a higher percentage of delayed flights?

Here is code to answer that question:

# Create a summary table of percent delayed by airline.
<!-- d %>% 
  group_by(airline) %>% 
  summarize(n = n(),
            percent_delay = (mean(delay_indicator) * 100) %>% round(1))  -->
Note that because delay_indicator is numeric (a binary 0/1 variable) calculating the mean of the vector returns the proportion of 1s, which, multiplied by 100, is equivalent to the percentage of delayed flights.

Write your own code to create a table summarizing the percentage of delayed flights by airline and route.

Notice that these tables—percent delayed by airline vs. percent delayed by airline and route— contain conflicting information. How should you answer the question of whether RegionEx has a higher percentage of delayed flights? Is the the COO correct? And, if not, why not?


```{r}

```


## Q4
Compare the scheduled flight durations for the two airlines on each of their four routes. Also compare the actual flight durations for the two airlines. What do you notice? If the two airlines had the same scheduled duration, what impact would this have on their delay records?


```{r}

```


## Q5
Does the data support the claim that the on‐time performance of RegionEx is worse than that of MDA? Write a paragraph in which you argue a position. In your answer, please incorporate quantitative evidence from the earlier questions.
```{r}

```

