---
title: "A1_Orton_Jamison"
output: 
  html_document: 
    toc: true
    toc_depth: 4
---

# Loading Packages

```{r loading packages}
library(tidyverse)
library(caret)
library(RWeka)
```

# Building first dataframe

```{r customer dataframe}
first_name <- c("John","Jane","Sam","Travis","Cindy")
last_name <- c("Smith","Dow","Alvares","Marks","Lou")
age <- c(21,52,33,16,41)
home_state <-c("AZ","UT","CA","MO","FL")
df <- data.frame(first_name,last_name,age,home_state)
df
nrow(df)
ncol(df)
dim(df)
```
## Observations

Something I observe in this dataframe is that there 5 rows (corresponding to 5 different customers) and 4 columns (corresponding to firstname, lastname, age, and state). There are 3 columns of the char type and only 1 of integer. The nice thing is that when you look at the dataframe all 5 rows have data in them so each vector has 5 items in them.

# Multiply_six Function

```{r first function (multiply_six)}
multiply_six <- function(a) {
  a * 6
}
```

## Using the function

```{r using multiply_six}
first_vector <- c(1,2,3,4,5,6)

second_vector <- multiply_six(first_vector)
second_vector

third_vector <- multiply_six(second_vector)
third_vector

fourth_vector <- multiply_six(third_vector)
fourth_vector

fifth_vector <- multiply_six(fourth_vector)
fifth_vector

sixth_vector <- multiply_six(fifth_vector)
sixth_vector
```